Neon: Nuclear Norm to Beat Muon

### 2025-05-13 poster/neon_poster.tex
Added mentors
changed kernel norm to nuclear norm in poster. There is no kernel norm, I am idiot

\begin{textblock}{23}(0,0.6)
{
\LARGE
Alexey Kravatskiy, Ivan Kozyrev, Nikolay Kozlov, Alexander Vinogradov
}

@@

\hrule\medskip
\Head{Introduction}\\
Recent advances in optimization techniques have highlighted the benefits of leveraging the matrix structure of neural network weights during training. Optimizers like Muon (Jordan et al.) and Shampoo (Gupta, Koren and Singer) have shown promise, but at a significantly higher computational cost than traditional methods like Adam. To bridge this gap, we propose Neon, a new optimizer that builds upon the framework of Bernstein and Newhouse. By using alternative norms, such as kernel norm (Kernel-Neon) or custom $F*$ norm ($F*$-Neon), we induce low-rank update matrices that enable more efficient computation. We evaluate the performance of Neon, Muon, and Adam on MLP, CIFAR10, and NanoGPT, and demonstrate [resting results here].

\medskip
\hrule\medskip
\Head{Neon's update rule}\\
Bernstein and Newhouse suggest obtaining the update step for a weight matrix $W$ as a solution to the optimization problem:
\begin{equation}
    \langle G, \delta W \rangle + \frac{\lambda}{2} \norm{\delta W}^2 \to \min_{\delta W}\,,
\end{equation}
where $G$ is a gradient-like matrix obtained via backpropagation. Setting norm to RMS-to-RMS norm (scaled version of Spectral norm) produces Muon. We consider two different choices instead:
\begin{enumerate}
    \item Choosing kernel norm, $\norm{\cdot}_*$, produces rank-1 update, defined by 
    \begin{equation}\label{eqn:update_star}
        \delta W = -\frac{1}{\lambda} u_1 \sigma_1 v_1^T\,,
    \end{equation}
    where $\sigma_1$ is largest singular value of $G$, and $u_1$, $v_1$ are corresponding singular values.

@@

\noindent\rule[-5pt]{.8\textwidth}{0.4pt}\\
{\bf Algorithm 1} Kernel-Neon update step for linear layer\\
\noindent\rule[10pt]{.8\textwidth}{0.4pt}
\vspace{-10pt}
\noindent\begin{tabbing}
    {\bf Input:} $\lambda$, gradient-like matrix $G$. \\
    {\bf Output:} Weight matrix update $\delta W$.\\*[\smallskipamount]
    \qquad \= 1.\ $U, \Sigma, V := \textrm{Lanczos}(G, 1)$. \\
    {\bf Return} $-\frac{1}{\lambda}U \Sigma V^T$.
\end{tabbing}
\vspace{-8pt}
\noindent\rule[10pt]{.8\textwidth}{0.4pt}

>>>>

\begin{textblock}{23}(0,0.6)
{
\LARGE
Alexey Kravatskiy, Ivan Kozyrev, Nikolay Kozlov, Alexander Vinogradov, Daniil Merkulov, Ivan Oseledets
}

@@

\hrule\medskip
\Head{Introduction}\\
Recent advances in optimization techniques have highlighted the benefits of leveraging the matrix structure of neural network weights during training. Optimizers like Muon (Jordan et al.) and Shampoo (Gupta, Koren and Singer) have shown promise, but at a significantly higher computational cost than traditional methods like Adam. To bridge this gap, we propose Neon, a new optimizer that builds upon the framework of Bernstein and Newhouse. By using alternative norms, such as nuclear norm (Nuclear-Neon) or custom $F*$ norm ($F*$-Neon), we induce low-rank update matrices that enable more efficient computation. We evaluate the performance of Neon, Muon, and Adam on MLP, CIFAR10, and NanoGPT, and demonstrate [resting results here].

\medskip
\hrule\medskip
\Head{Neon's update rule}\\
Bernstein and Newhouse suggest obtaining the update step for a weight matrix $W$ as a solution to the optimization problem:
\begin{equation}
    \langle G, \delta W \rangle + \frac{\lambda}{2} \norm{\delta W}^2 \to \min_{\delta W}\,,
\end{equation}
where $G$ is a gradient-like matrix obtained via backpropagation. Setting norm to RMS-to-RMS norm (scaled version of Spectral norm) produces Muon. We consider two different choices instead:
\begin{enumerate}
    \item Choosing nuclear norm, $\norm{\cdot}_*$, produces rank-1 update, defined by 
    \begin{equation}\label{eqn:update_star}
        \delta W = -\frac{1}{\lambda} u_1 \sigma_1 v_1^T\,,
    \end{equation}
    where $\sigma_1$ is largest singular value of $G$, and $u_1$, $v_1$ are corresponding singular values.

@@

\noindent\rule[-5pt]{.8\textwidth}{0.4pt}\\
{\bf Algorithm 1} Nuclear-Neon update step for linear layer\\
\noindent\rule[10pt]{.8\textwidth}{0.4pt}
\vspace{-10pt}
\noindent\begin{tabbing}
    {\bf Input:} $\lambda$, gradient-like matrix $G$. \\
    {\bf Output:} Weight matrix update $\delta W$.\\*[\smallskipamount]
    \qquad \= 1.\ $U, \Sigma, V := \textrm{Lanczos}(G, 1)$. \\
    {\bf Return} $-\frac{1}{\lambda}U \Sigma V^T$.
\end{tabbing}
\vspace{-8pt}
\noindent\rule[10pt]{.8\textwidth}{0.4pt}
